NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
Edits to the conclusions

Calcium binding affinity to membranes containing the negatively charged PS lipids can be
experimentally quantified by measuring the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the measurement is not compromised by the dehydrated lipid-ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of an ion-free reference state
in the presence of negatively charged lipids, our simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters upon increasing Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
indicating too strong calcium binding affinity.
The only exception is the CHARMM36 simulation utilizing the NBfix
interaction for calcium~\cite{kim16}, wherein the modulation of order parameters is underestimated, 
indicating weaker binding affinity than in experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
is underestimated in CHARMM36 simulations when using the NBfix for calcium \cite{kim16}, but overestimated 
in all the other tested models. This is evident in the calcium density distributions where almost all Ca$^{2+}$ ions bind to the membrane interface in all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong dependence of CaCl$_2$ with small concentrations that saturates below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
tested simulation models, including CHARMM36 which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behaviour
for the two POPS $\alpha$-carbon order parameters with added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
This is in contrast to the PC headgroup, where
qualitatively correct reponse to bound ions is observed
in all simulation models, despite significant discrepancies in the headgroup
structures produced in salt-free simulations~\cite{catte16}. Therefore, we conclude that
improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.
\section{Conclusions}
Lipids with PS headgroups, and their interactions with ions,
play an important role in lipid-mediated signaling
processes \cite{leventis10,yeung08}. Attemps to use MD
simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PS
headgroups \cite{melcrova16,valentine18,hallock18}.  
Here (as was previously done for PC lipids \cite{botan15,catte16})
we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
of the headgroup structure and ion binding affinity to PS lipids
in available MD force fields.
The main advantage of this approach is the direct connection
between the accurately measured experimental order parameters and the simulations,
which reduces the ambiguity in the interpretation of experiments.

First, we complemented the available experimental data \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters.
Comparison to these data revealed that none of the available force fields
%tested using the NMRlipids open collaboration
reproduce the PS headgroup order parameters within the experimental accuracy. However,
the best models for the serine headgroup suggested a characteristic rigid conformation for its
carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
concentrations \cite{roux90} then showed that the tested MD force fields
overestimate the cation binding affinity to these bilayers with two exceptions. 1) The MacRog simulation with potassium counterion appear to produce the most realistic monovalent ion binding
affinity to PS-containing lipid bilayers. 2) The CHARMM36 force field with the recently introduced
NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
The experimentally measured response of the PS headgroup order parameter to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
qualitatively reproduced in any of the tested force fields, which underlines the need for more accurate
the MD force fields to study interactions between PS lipids and other biomolecules.
This is in contrast to the previous results with PC lipids,
where the experimentally measured headgroup order parameter responses to the bound charge
were in qualitative agreement even though the headgroup structures themselves were
incorrect and the cation binding affinities overestimated \cite{catte16}.

Our results pave the way for the development of better MD force fields
for PS lipids. Using the headgroup order parameters, we were able to evaluate the
quality of various conformational ensembles in different force fields.
This can guide the development of force fields that would correctly reproduce the
conformations sampled by PS headgroups. The experimental dataset of headgroup order
parameters from POPC:POPS (5:1) mixture with different cation concentrations can
be used to improve cation binding details in the force fields, as recently demonstrated
for POPC using the electronic continuum correction \cite{melcr18}.
Similar study for POPS is being progressed separately \cite{ECCpops}.

>>>>

Calcium binding affinity to membranes containing the negatively charged PS lipids can be
experimentally measured by detecting the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the measurement is not complicated by the dehydrated lipid-ion complexes and phase separation and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of an ion-free reference state
in the presence of negatively charged lipids, our simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
in POPC:POPS (5:1) mixtures as Ca$^{2+}$ concentration increases (Fig. \ref{changesWITHCaClPS}),
indicating overestimated calcium binding affinity.
The only exception is the CHARMM36 model with the NBfix
interaction employed for calcium~\cite{kim16}, which underestimates the modulation of order parameters, 
indicating weaker binding affinity than experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
is underestimated in CHARMM36 simulations when using the NBfix for calcium \cite{kim16}, but overestimated 
in all the other tested models. This is evident in the calcium density distributions where almost all Ca$^{2+}$ ions bind to the membrane interface in all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
the smaller $\alpha$-carbon. All these changes are significantly overestimated in the
tested simulation models, including CHARMM36 which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behaviour
for the POPS $\alpha$-carbon order parameters with added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
This is in contrast to the PC headgroup, where
qualitatively correct reponse to bound ions is observed
in all simulation models, despite significant discrepancies in the headgroup
structure without additional ions~\cite{catte16}. Therefore, we conclude that
improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.
\section{Conclusions}

%Lipids with PS headgroups, and their interactions with ions, play an important role in lipid-mediated signaling processes \cite{leventis10,yeung08}. Attemps to use MD simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PSheadgroups \cite{melcrova16,valentine18,hallock18}.  
%Here (as was previously done for PC lipids \cite{botan15,catte16})
Here, we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
of the headgroup structure and the ion binding affinity 
in available PS MD models.
The main advantage of this approach is the direct connection
between the accurately measurable experimental order parameters and the simulations,
which reduces the ambiguity in the interpretation of experiments.

First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters, and the proceeded to compare MD simulation results from several force fields to the experimental data
This revealed that none of the force fields
%tested using the NMRlipids open collaboration
reproduce the PS headgroup order parameters within the experimental accuracy. However,
the best models for the serine headgroup suggested a characteristic rigid conformation for its
carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
concentrations \cite{roux90} then showed that the tested MD force fields
overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
The experimentally measured response of the PS headgroup order parameter to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
qualitatively reproduced in any of the tested force fields, which underlines the need for more accurate
the MD force fields to study the biological functioning of PS lipids.
This is in contrast to the previous results with PC lipids,
where the experimentally measured headgroup order parameter responses to the bound charge
were in qualitative agreement even though the headgroup structures themselves were
incorrect and the cation binding affinities overestimated \cite{catte16}.

We expect our results to pave the way for the development of better MD force fields
for PS lipids. The quality of conformational ensembles produced by the various models can be evaluated based on the headgroup order parameters, which we hope will to guide the development of force fields towards models that correctly reproduce the PS headgroup structures. The cation binding in the force fields can be improved based on the experimental headgroup order parameter data from POPC:POPS (5:1) mixtures under different cation concentrations. This was already demonstrated for POPC using the electronic continuum correction \cite{melcr18}, and a
similar study for POPS is being progressed separately \cite{ECCpops}.
